---
title: "Convex Hull Theorem Illustration"
subtitle: "Figures 1-4"
execute:
  warning: false
  message: false
  cache: false
---

## A convex hull with 100 points (Figure 1)


```{r example}
# Load required libraries
library(ggplot2)
library(grDevices)

# Set seed for reproducibility
set.seed(50720235)

# Generate random uniform points in rectangle [0,4] by [0,2]
n_points <- 100  # Number of random points
x <- runif(n_points, min = 0, max = 4)
y <- runif(n_points, min = 0, max = 2)

# Create data frame
points_df <- data.frame(x = x, y = y)

# Compute convex hull
hull_indices <- chull(x, y)
hull_points <- points_df[hull_indices, ]

# Add first point at the end to close the polygon
hull_points <- rbind(hull_points, hull_points[1, ])

# Count extreme points (vertices of convex hull)
n_extreme_points <- length(hull_indices)

# Create the plot
p <- ggplot() +
  # Add shaded convex hull region in light blue
  geom_polygon(data = hull_points, 
               aes(x = x, y = y), 
               fill = "lightblue", 
               alpha = 0.6, 
               color = "blue", 
               linewidth = 1) +
  
  # Add all random points in gray
  geom_point(data = points_df, 
             aes(x = x, y = y), 
             color = "gray50", 
             size = 2) +
  
  # Add extreme points (hull vertices) in red
  geom_point(data = points_df[hull_indices, ], 
             aes(x = x, y = y), 
             color = "red", 
             size = 3) +
  
  # Formatting
  xlim(0, 4) +
  ylim(0, 2) +
  labs(title = paste("Convex Hull of", n_points, "Random Points"),
       subtitle = paste("Number of extreme points:", n_extreme_points),
       x = "X coordinate", 
       y = "Y coordinate") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank(),
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5)) +
  coord_fixed()  # Maintain aspect ratio

# Display the plot
print(p)

# Print summary information
cat("Summary:\n")
cat("Total points generated:", n_points, "\n")
cat("Number of extreme points (convex hull vertices):", n_extreme_points, "\n")
cat("Proportion of extreme points:", round(n_extreme_points/n_points, 3), "\n")


# Save the plot file (uncomment as necessary)
# ggsave("convex_hull_plot.png", plot = p, width = 8, height = 6, units = "in")
```


## Code for Figures 2 and 3 that illustrate the theorem for different distributions 

This shows the generation of the figures for the convex hull verifications in section 2.2.
Added on is the case of uniform points on a disk.

```{r setup}
# Load required libraries
library(ggplot2)
library(dplyr)
library(gridExtra)
library(geometry)
library(MASS)  # for mvrnorm
library(sp)    # for point.in.polygon
```

```{r uniformdisk, eval=FALSE}
# Generate points from uniform distribution on a disk
generate_uniform_disk_points <- function(n, d, radius = 1) {
  if (d != 2) {
    stop("Uniform disk distribution is only implemented for d = 2")
  }
  
  # Generate angles uniformly between 0 and 2π
  theta <- runif(n, 0, 2*pi)
  
  # Generate radii with r ~ sqrt(U) for uniformity
  # (in 2D, need r ~ U^(1/2) for uniform distribution in disk)
  r <- radius * sqrt(runif(n))
  
  # Convert to Cartesian coordinates
  x <- r * cos(theta)
  y <- r * sin(theta)
  
  return(cbind(x, y))
}
```

## Generation functions for different distributions

```{r distributions}
#| code-fold: TRUE
# Function to generate points from different distributions
generate_distribution_points <- function(n, d, distribution, correlation = 0.5) {
  if(distribution == "uniform_rectangle") {
    # Uniform in [0,4] x [0,2] (only for d=2) or [0,4] x [0,2] x [0,1] (for d=3)
    if(d == 2) {
      X <- cbind(runif(n, 0, 4), runif(n, 0, 2))
    } else if(d == 3) {
      X <- cbind(runif(n, 0, 4), runif(n, 0, 2), runif(n, 0, 1))
    }
  } else if (distribution == "uniform_disk") {
   # Uniform distribution on disk
  if(d == 2) {
    # 2D disk implementation
    theta <- runif(n, 0, 2*pi)
    r <- sqrt(runif(n))
    x <- r * cos(theta)
    y <- r * sin(theta)
    return(cbind(x, y))
  } else if(d == 3) {
    # 3D ball implementation
    # Generate points from a 3D ball using spherical coordinates
    theta <- runif(n, 0, 2*pi)      # Azimuthal angle
    phi <- acos(2*runif(n) - 1)     # Polar angle
    r <- runif(n)^(1/3)             # Radius (cube root for uniform distribution)
    
    x <- r * sin(phi) * cos(theta)
    y <- r * sin(phi) * sin(theta)
    z <- r * cos(phi)
    
    return(cbind(x, y, z))
  }  }   else if(distribution == "gaussian_independent") {
    # Independent standard Gaussians
    X <- matrix(rnorm(n * d, 0, 1), ncol = d)
  } else if(distribution == "gaussian_correlated") {
    # Correlated Gaussians with correlation r
    Sigma <- matrix(correlation, nrow = d, ncol = d)
    diag(Sigma) <- 1
    X <- mvrnorm(n, mu = rep(0, d), Sigma = Sigma)
  }
  return(X)
}
```

## Estimate Dn by Monte Carlo

```{r MCDn}
#| code-fold: TRUE
# Function to estimate D_n using Monte Carlo
estimate_D_n <- function(hull_points, distribution, d, correlation = 0.5, n_test = 5000) {
  # Generate test points from the same distribution
  test_points <- generate_distribution_points(n_test, d, distribution, correlation)
  
  if(d == 2) {
    # Use point.in.polygon for 2D
    if(nrow(hull_points) < 3) return(1)  # If no proper hull, D_n = 1
    
    # Ensure hull is closed
    if(!all(hull_points[1,] == hull_points[nrow(hull_points),])) {
      hull_points <- rbind(hull_points, hull_points[1,])
    }
    
    inside_count <- sum(point.in.polygon(test_points[,1], test_points[,2], 
                                        hull_points[,1], hull_points[,2]) > 0)
    D_n <- 1 - inside_count / n_test
    
  } else if(d == 3) {
    # Simple approximation for 3D using bounding box
    if(nrow(hull_points) < 4) return(1)
    
    bbox_min <- apply(hull_points, 2, min)
    bbox_max <- apply(hull_points, 2, max)
    
    # Count points inside bounding box (conservative approximation)
    inside_bbox <- apply(test_points, 1, function(pt) {
      all(pt >= bbox_min) && all(pt <= bbox_max)
    })
    
    # For Gaussian distributions, estimate the total "relevant" measure
    if(distribution == "uniform_rectangle") {
      total_measure <- 4 * 2 * 1  # [0,4] x [0,2] x [0,1]
      inside_count <- sum(inside_bbox)
      D_n <- 1 - inside_count / n_test
    } else if(distribution == "uniform_disk") {
      # For uniform disk, same approach as above
      inside_count <- sum(inside_bbox)
      D_n <- 1 - inside_count / n_test
    }  else {
      # For Gaussians, use a more conservative estimate
      # Assume most probability mass is within some reasonable bounds
      center <- colMeans(hull_points)
      inside_count <- sum(apply(test_points, 1, function(pt) {
        sqrt(sum((pt - center)^2)) <= 0.6 * sqrt(sum((bbox_max - bbox_min)^2))
      }))
      D_n <- 1 - inside_count / n_test
    }
  }
  
  return(max(0, min(1, D_n)))  # Clamp to [0,1]
}
```

## Illustration of the main theorem

```{r mainthm}
#| code-fold: TRUE
# Main simulation function for the theorem
simulate_main_theorem <- function(n, d, distribution, correlation = 0.5, n_simulations = 300) {
  results <- data.frame(
    simulation = integer(n_simulations),
    V_n = integer(n_simulations),
    D_n = numeric(n_simulations),
    D_n_minus_1 = numeric(n_simulations),
    V_n_over_n = numeric(n_simulations),
    difference_squared = numeric(n_simulations),
    D_diff = numeric(n_simulations),
    n = integer(n_simulations),
    d = integer(n_simulations),
    distribution = character(n_simulations),
    stringsAsFactors = FALSE
  )
  
  for(i in 1:n_simulations) {
    # Generate n points
    X_n <- generate_distribution_points(n, d, distribution, correlation)
    
    # Generate n-1 points for D_{n-1}
    X_n_minus_1 <- X_n[1:(n-1), ]
    
    # Compute convex hulls
    if(d == 2) {
      # For n points
      hull_indices_n <- chull(X_n[,1], X_n[,2])
      V_n <- length(hull_indices_n)
      hull_points_n <- X_n[hull_indices_n, ]
      
      # For n-1 points
      hull_indices_n_minus_1 <- chull(X_n_minus_1[,1], X_n_minus_1[,2])
      hull_points_n_minus_1 <- X_n_minus_1[hull_indices_n_minus_1, ]
      
    } else if(d == 3) {
      # For n points
      hull_faces_n <- convhulln(X_n, options = "Pp")
      V_n <- length(unique(as.vector(hull_faces_n)))
      hull_vertices_n <- X_n[unique(as.vector(hull_faces_n)), ]
      
      # For n-1 points
      hull_faces_n_minus_1 <- convhulln(X_n_minus_1, options = "Pp")
      hull_vertices_n_minus_1 <- X_n_minus_1[unique(as.vector(hull_faces_n_minus_1)), ]
    }
    
    # Estimate D_n and D_{n-1}
    if(d == 2) {
      D_n <- estimate_D_n(hull_points_n, distribution, d, correlation)
      D_n_minus_1 <- estimate_D_n(hull_points_n_minus_1, distribution, d, correlation)
    } else {
      D_n <- estimate_D_n(hull_vertices_n, distribution, d, correlation)
      D_n_minus_1 <- estimate_D_n(hull_vertices_n_minus_1, distribution, d, correlation)
    }
    
    # Compute the quantities in the theorem
    V_n_over_n <- V_n / n
    difference_squared <- (V_n_over_n - D_n_minus_1)^2
    D_diff <- abs(D_n - D_n_minus_1)
    
    results[i, ] <- list(i, V_n, D_n, D_n_minus_1, V_n_over_n, 
                        difference_squared, D_diff, n, d, distribution)
    
    #if(i %% 50 == 0) cat("Completed", i, "of", n_simulations, "simulations\n")
  }
  
  return(results)
}
```

## Run simulations for sample sizes from 50 to 600.

```{r simulations}
#| code-fold: true
# Set seed for reproducibility
set.seed(20421)

# Simulation parameters
sample_sizes <- c(50, 100, 200, 400, 600)
dimensions <- c(2, 3)
distributions <- c("uniform_rectangle", "uniform_disk", "gaussian_independent", "gaussian_correlated")
correlation <- 0.8
n_sims <- 250

#cat("Running main theorem simulations...\n")

# Run all simulations
all_theorem_data <- data.frame()

for(d in dimensions) {
  for(dist in distributions) {
    for(n in sample_sizes) {
      #cat("\n--- Running", dist, "d =", d, "n =", n, "---\n")
      
      sim_data <- simulate_main_theorem(n, d, dist, correlation, n_sims)
      all_theorem_data <- rbind(all_theorem_data, sim_data)
    }
  }
}

# Clean up distribution labels
all_theorem_data$distribution_clean <- case_when(
  all_theorem_data$distribution == "uniform_rectangle" ~ "Uniform Rectangle",
  all_theorem_data$distribution == "uniform_disk" ~ "Uniform Disk",
  all_theorem_data$distribution == "gaussian_independent" ~ "Independent Gaussian",
  all_theorem_data$distribution == "gaussian_correlated" ~ paste0("Corr. Gaussian (r=", correlation, ")"),
  TRUE ~ all_theorem_data$distribution
)

cat("\nSimulations complete. Creating plots...\n")

# Compute summary statistics
theorem_summary <- all_theorem_data %>%
  group_by(distribution_clean, d, n) %>%
  summarise(
    mean_squared_diff = mean(difference_squared),
    theoretical_bound = (8*d + 9) / n,
    mean_D_diff = mean(D_diff),
    theoretical_D_bound = (d + 1) / n,
    mean_V_n_over_n = mean(V_n_over_n),
    mean_D_n_minus_1 = mean(D_n_minus_1),
    sd_squared_diff = sd(difference_squared),
    .groups = 'drop'
  )
```

## Creation of Figure 2

```{r plot2}

# FIGURE 2: Main theorem bound - MSE
fig2 <- ggplot(theorem_summary, aes(x = n)) +
  geom_line(aes(y = mean_squared_diff, color = distribution_clean), size = 1) +
  geom_line(aes(y = theoretical_bound), linetype = "dashed", color = "black", size = 1) +
  facet_wrap(~paste("d =", d), scales = "free_y") +
  labs(x = "Sample size (n)",
       y = "Mean squared error",
       color = "Distribution") +
  theme_minimal() +
  theme(legend.position = c(0.15, 0.35),
        legend.justification = c("left","bottom"))+
  coord_cartesian(xlim = c(100, max(theorem_summary$n)))+
  scale_y_log10() 
  #+
  #+annotate("text", x = 400, y = 0.1, label = "Theoretical bound", 
  #         color = "black", size = 3)

print(fig2)
# ggsave("main_theorem_mse_bound.png", fig2, width = 7, height = 4,dpi = 300, bg = "white")
```

## Second part of theorem, Figure showing D_n difference

```{r}
# FIGURE 3: Second part of theorem - D_n difference
fig3 <- ggplot(theorem_summary, aes(x = n)) +
  geom_line(aes(y = mean_D_diff, color = distribution_clean), linewidth = 1) +
  geom_line(aes(y = theoretical_D_bound), linetype = "dashed", color = "black", size = 1) +
  facet_wrap(~paste("d =", d), scales = "free_y") +
  labs(x = "Sample size (n)",
       y = "Expected value of the differences in Ds",
       color = "Distribution") +
  theme_minimal() +
  theme(legend.position = c(0.17,0.45),
        legend.justification = c("left","bottom"))+
  coord_cartesian(xlim = c(100, max(theorem_summary$n)))+
  scale_y_log10() #+
 # annotate("text", x = 400, y = 0.01, label = "Theoretical bound", 
 #           color = "black", size = 3)

print(fig3)
# To save to a file
# ggsave("main_theorem_D_diff_bound.png", fig3, 
  #       width = 7, height = 4, dpi = 300, bg = "white")
```

## Convergence of expectations 

```{r}
# FIGURE 4: Convergence of expectations
fig4 <- ggplot(theorem_summary, aes(x = n)) +
  geom_line(aes(y = mean_V_n_over_n, color = paste(distribution_clean, expression(-V[n]/n))), 
            size = 1.5, alpha = 0.7) +
  geom_line(aes(y = mean_D_n_minus_1, color = paste(distribution_clean, expression(-D[n-1])
)), 
            size = 1.5, alpha = 0.7) +
  facet_wrap(~paste("d =", d), scales = "free_y") +
  labs(#title = expression("Convergence: " * E[V[n]/n] * " and " * E[D[n-1]]),
       #subtitle = "The theorem states these should be equal in expectation",
       x = "Sample size (n)",
       y = "Expected value",
       color = "Quantity") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  guides(color = guide_legend(ncol = 2))

print(fig4)
#ggsave("main_theorem_convergence.png", fig4, 
#        width = 6, height = 4, dpi = 300, bg = "white")


# Print summary statistics
cat("\n=== MAIN THEOREM VERIFICATION ===\n")
print(theorem_summary)

cat("\nBound violations for MSE:\n")
mse_violations <- theorem_summary[theorem_summary$mean_squared_diff > theorem_summary$theoretical_bound, ]
if(nrow(mse_violations) > 0) {
  print(mse_violations[, c("distribution_clean", "d", "n", "mean_squared_diff", "theoretical_bound")])
} else {
  cat("None! The MSE bound holds in all cases.\n")
}

cat("\nBound violations for D difference:\n")
D_violations <- theorem_summary[theorem_summary$mean_D_diff > theorem_summary$theoretical_D_bound, ]
if(nrow(D_violations) > 0) {
  print(D_violations[, c("distribution_clean", "d", "n", "mean_D_diff", "theoretical_D_bound")])
} else {
  cat("None! The D difference bound holds in all cases.\n")
}


cat("\nExpectation equality check (should be close to 0):\n")
expectation_diff <- theorem_summary %>%
  mutate(expectation_difference = abs(mean_V_n_over_n - mean_D_n_minus_1)) %>%
  arrange(desc(expectation_difference))

# Print just the relevant columns
print(head(expectation_diff[, c("distribution_clean", "d", "n", "expectation_difference")], 10))


cat("\n=== SUMMARY ===\n")
cat("This code illustrates the main theorem with:\n")
cat("1. MSE bound: E[(V_n/n - D_{n-1})^2] ≤ (8d+10)/n\n")
cat("2. D difference bound: E[|D_n - D_{n-1}|] ≤ (d+1)/n\n")
cat("3. Expectation equality: E[V_n/n] = E[D_{n-1}]\n")
cat("4. Three distributions across 2D and 3D\n")

```



## Corollary Illustration with creation of Fig 4

```{r}
# Load required libraries
library(ggplot2)
library(dplyr)
library(gridExtra)
library(geometry)

# Function to simulate the corollary bound - with 4-simplex added
simulate_corollary_with_simplex <- function(n, d, K_type = "unit_cube", n_simulations = 400) {
  
  # Define the convex set K and its true volume
  if(K_type == "unit_cube") {
    vol_K_true <- 1
    generate_points <- function(n) matrix(runif(n * d, 0, 1), ncol = d)
  } else if(K_type == "unit_ball") {
    if(d == 2) {
      vol_K_true <- pi
      generate_points <- function(n) {
        points <- matrix(0, nrow = n, ncol = 2)
        count <- 0
        while(count < n) {
          candidate <- runif(2, -1, 1)
          if(sum(candidate^2) <= 1) {
            count <- count + 1
            points[count, ] <- candidate
          }
        }
        return(points)
      }
    } else if(d == 3) {
      vol_K_true <- 4*pi/3
      generate_points <- function(n) {
        points <- matrix(0, nrow = n, ncol = 3)
        count <- 0
        while(count < n) {
          candidate <- runif(3, -1, 1)
          if(sum(candidate^2) <= 1) {
            count <- count + 1
            points[count, ] <- candidate
          }
        }
        return(points)
      }
    }
  } else if(K_type == "simplex") {
    if(d == 2) {
      # Triangle: vertices (0,0), (1,0), (0,1)
      vol_K_true <- 0.5
      generate_points <- function(n) {
        u1 <- runif(n)
        u2 <- runif(n)
        x <- ifelse(u1 + u2 <= 1, u1, 1 - u1)
        y <- ifelse(u1 + u2 <= 1, u2, 1 - u2)
        return(cbind(x, y))
      }
    } else if(d == 3) {
      # Regular tetrahedron (4-simplex in 3D)
      # Vertices: (0,0,0), (1,0,0), (0,1,0), (0,0,1)
      vol_K_true <- 1/6  # Volume of tetrahedron with these vertices
      generate_points <- function(n) {
        # Generate uniform points in 3-simplex using Dirichlet method
        # Generate 4 exponential random variables, normalize to get Dirichlet
        points <- matrix(0, nrow = n, ncol = 3)
        for(i in 1:n) {
          # Generate 4 exponential(1) random variables
          exp_vars <- rexp(4, rate = 1)
          # Normalize to get Dirichlet(1,1,1,1) - uniform on 3-simplex
          dirichlet_vars <- exp_vars / sum(exp_vars)
          # The first 3 coordinates give us a point in the tetrahedron
          # (the 4th coordinate is implicit: 1 - sum of first 3)
          points[i, ] <- dirichlet_vars[1:3]
        }
        return(points)
      }
    }
  }
  
  # Storage for individual simulation results
  lhs_values <- numeric(n_simulations)
  vol_ratios <- numeric(n_simulations)
  vertex_terms <- numeric(n_simulations)
  V_n_values <- integer(n_simulations)
  
  for(i in 1:n_simulations) {
    # Step 1: Generate n random points in K
    X <- generate_points(n)
    
    # Step 2: Compute convex hull and V_n
    if(d == 2) {
      hull_indices <- chull(X[,1], X[,2])
      V_n <- length(hull_indices)
      vol_convhull <- convhulln(X, options = "FA")$vol
    } else if(d == 3) {
      hull_faces <- convhulln(X, options = "Pp")
      V_n <- length(unique(as.vector(hull_faces)))
      vol_convhull <- convhulln(X, options = "FA")$vol
    }
    
    # Step 3: Compute the volume estimator
    if(V_n < n) {  # Avoid division by zero
      vol_K_hat <- vol_convhull / (1 - V_n/n)
    } else {
      lhs_values[i] <- NA
      vol_ratios[i] <- NA
      vertex_terms[i] <- NA
      V_n_values[i] <- V_n
      next
    }
    
    # Step 4: Compute the components of the LHS
    vertex_component <- (n - V_n)^2 / ((8*d + 10) * n)
    volume_component <- (vol_K_hat / vol_K_true - 1)^2
    
    # Step 5: Compute the full LHS for this simulation
    lhs_values[i] <- vertex_component * volume_component
    vol_ratios[i] <- vol_K_hat / vol_K_true
    vertex_terms[i] <- vertex_component
    V_n_values[i] <- V_n
    
    #if(i %% 100 == 0) cat("Simulation", i, "of", n_simulations, "completed\n")
  }
  
  # Remove NA values
  valid_indices <- !is.na(lhs_values) & is.finite(lhs_values)
  
  # Step 6: Compute the empirical expectation
  empirical_expectation <- mean(lhs_values[valid_indices])
  
  return(list(
    n = n,
    d = d,
    K_type = K_type,
    empirical_expectation = empirical_expectation,
    individual_lhs = lhs_values[valid_indices],
    vol_ratios = vol_ratios[valid_indices],
    vertex_terms = vertex_terms[valid_indices],
    V_n_values = V_n_values[valid_indices],
    n_valid_sims = sum(valid_indices),
    theoretical_bound = 1
  ))
}

# Set seed for reproducibility
set.seed(42)

# Simulation parameters - simplified to focus on the nice cases
sample_sizes <- c(50, 100, 200, 400, 600)
dimensions <- c(2, 3)
# Now using simplex for both dimensions: triangle in 2D, tetrahedron in 3D
K_types <- c("unit_cube", "unit_ball", "simplex")
n_sims <- 500

# cat("Running corollary simulations with 4-simplex...\n")

# Run all simulations and collect results
results_list <- list()
summary_data <- data.frame()

for(d in dimensions) {
  for(K_type in K_types) {
    for(n in sample_sizes) {
      #cat("\n--- Running for", K_type, "d =", d, "n =", n, "---\n")
      
      result <- simulate_corollary_with_simplex(n, d, K_type, n_sims)
      results_list[[length(results_list) + 1]] <- result
      
      # Add to summary
      summary_data <- rbind(summary_data, data.frame(
        K_type = K_type,
        d = d,
        n = n,
        empirical_expectation = result$empirical_expectation,
        median_lhs = median(result$individual_lhs),
        q95_lhs = quantile(result$individual_lhs, 0.95),
        max_lhs = max(result$individual_lhs),
        prop_exceed_1 = mean(result$individual_lhs > 1),
        mean_vol_ratio = mean(result$vol_ratios),
        n_valid_sims = result$n_valid_sims
      ))
    }
  }
}

# Clean up labels
summary_data$K_type_clean <- case_when(
  summary_data$K_type == "unit_cube" ~ "Unit Cube",
  summary_data$K_type == "unit_ball" ~ "Unit Ball",
  summary_data$K_type == "simplex" & summary_data$d == 2 ~ "Triangle",
  summary_data$K_type == "simplex" & summary_data$d == 3 ~ "Tetrahedron",
  TRUE ~ summary_data$K_type
)

cat("\nSimulations complete. Creating plots...\n")

# FIGURE 1: The main result - empirical expectation vs theoretical bound
fig1 <- ggplot(summary_data, aes(x = n, y = empirical_expectation, color = K_type_clean)) +
  geom_line(size = 1.5) +
  geom_point(size = 2.5) +
  geom_hline(yintercept = 1, color = "red", linetype = "dashed", size = 1.5) +
  facet_wrap(~paste("d =", d), scales = "free_y") +
  labs(title = "Corollary Verification: Empirical Expectation vs Theoretical Bound",
       subtitle = expression("Bound: " * E[frac((n-V[n])^2, (8*d+10)*n) * (frac(hat(vol)(K), vol(K)) - 1)^2] <= 1),
       x = "Sample size (n)",
       y = "Empirical expectation of LHS",
       color = "Convex set K") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        legend.position = "bottom") +
  scale_color_manual(values = c("Unit Cube" = "blue", "Unit Ball" = "red", 
                               "Triangle" = "green", "Tetrahedron" = "purple")) +
  annotate("text", x = 500, y = 0.9, label = "Theoretical bound = 1", 
           color = "red", size = 4)

# print(fig1)
# ggsave("corollary_with_simplex_main.pdf", fig1, width = 12, height = 6)

# FIGURE 4: Volume estimator performance comparison
vol_estimator_data <- data.frame()
for(i in 1:length(results_list)) {
  result <- results_list[[i]]
  vol_estimator_data <- rbind(vol_estimator_data, data.frame(
    K_type = result$K_type,
    d = result$d,
    n = result$n,
    mean_vol_ratio = mean(result$vol_ratios),
    sd_vol_ratio = sd(result$vol_ratios)
  ))
}

vol_estimator_data$K_type_clean <- case_when(
  vol_estimator_data$K_type == "unit_cube" ~ "Unit Cube",
  vol_estimator_data$K_type == "unit_ball" ~ "Unit Ball",
  vol_estimator_data$K_type == "simplex" & vol_estimator_data$d == 2 ~ "Triangle",
  vol_estimator_data$K_type == "simplex" & vol_estimator_data$d == 3 ~ "Tetrahedron",
  TRUE ~ vol_estimator_data$K_type
)

fig4 <- ggplot(vol_estimator_data, aes(x = n, y = mean_vol_ratio, color = K_type_clean)) +
  geom_line(size = 1.5) +
  geom_point(size = 2) +
  geom_ribbon(aes(ymin = mean_vol_ratio - 1.96*sd_vol_ratio/sqrt(n_sims), 
                  ymax = mean_vol_ratio + 1.96*sd_vol_ratio/sqrt(n_sims),
                  fill = K_type_clean), alpha = 0.3) +
  geom_hline(yintercept = 1, linetype = "dotted", color = "black", size = 1) +
  facet_wrap(~paste("d =", d), scales = "free_y") +
  labs(#title = "Volume Estimator Performance",
       #subtitle = expression("Bias of " * hat(vol)(K) * " = " * frac(vol(conv(X[1],...,X[n])), 1 - V[n]/n)),
       x = "Sample size (n)",
       y = "Ratio expected value to true value of vol(K)",
       color = "Convex set K",
       fill = "Convex set K") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        legend.position = "bottom") +
  scale_color_manual(values = c("Unit Cube" = "blue", "Unit Ball" = "red", 
                               "Triangle" = "green", "Tetrahedron" = "purple")) +
  scale_fill_manual(values = c("Unit Cube" = "blue", "Unit Ball" = "red", 
                              "Triangle" = "green", "Tetrahedron" = "purple"))

print(fig4)
# ggsave("volume_estimator_with_simplex.pdf", fig4, width = 6, height = 4)

# Detailed comparison - log scale to see the small values better

figsimplex <- ggplot(summary_data, aes(x = n, y = empirical_expectation, color = K_type_clean)) +
  geom_line(size = 1.5) +
  geom_point(size = 2.5) +
  geom_hline(yintercept = 1, color = "red", linetype = "dashed", size = 1.5) +
  facet_wrap(~paste("d =", d), scales = "free_y") +
  labs(title = "Corollary Bound: Log Scale View",
       subtitle = "Detailed view of how far below 1 the empirical expectations are",
       x = "Sample size (n)",
       y = "Empirical expectation of LHS (log scale)",
       color = "Convex set K") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        legend.position = "bottom") +
  scale_color_manual(values = c("Unit Cube" = "blue", "Unit Ball" = "red", 
                               "Triangle" = "green", "Tetrahedron" = "purple")) +
  scale_y_log10() +
  annotate("text", x = 500, y = 1, label = "Theoretical bound = 1", 
           color = "red", size = 4)

print(figsimplex)
#ggsave("corollary_log_scale_view.pdf", figsimplex, width = 8, height = 4)

```

## Supplementary plot of theoretical versus estimates

```{r printresults, eval=FALSE}
# Print summary results
cat("\n=== COROLLARY VERIFICATION WITH SIMPLEX ===\n")
cat("Theoretical bound: E[LHS] ≤ 1\n\n")

summary_table <- summary_data %>%
  select(K_type_clean, d, n, empirical_expectation, prop_exceed_1) %>%
  arrange(d, K_type_clean, n)

print(summary_table)

cat("\nMaximum empirical expectation by case:\n")
max_by_case <- summary_data %>%
  group_by(K_type_clean, d) %>%
  summarise(
    max_empirical_exp = max(empirical_expectation),
    min_n_for_max = n[which.max(empirical_expectation)],
    avg_empirical_exp = mean(empirical_expectation),
    .groups = 'drop'
  )
print(max_by_case)

cat("\nVolume estimator performance (bias from 1):\n")
vol_bias <- vol_estimator_data %>%
  group_by(K_type_clean, d) %>%
  summarise(
    avg_bias = mean(mean_vol_ratio - 1),
    max_bias = max(abs(mean_vol_ratio - 1)),
    .groups = 'drop'
  )
print(vol_bias)

cat("\nCases where empirical expectation > 1:\n")
violations <- summary_data[summary_data$empirical_expectation > 1, ]
if(nrow(violations) > 0) {
  print(violations[, c("K_type_clean", "d", "n", "empirical_expectation")])
} else {
  cat("None! The bound holds in all cases.\n")
}

cat("\n=== SUMMARY ===\n")
cat("All empirical expectations are smaller than 1, suggesting the bound is quite conservative.\n")
cat("The tetrahedron (4-simplex in 3D) case has been added.\n")
cat("Volume estimator shows good performance across all convex sets.\n")
```


The tetrahedron case uses the standard 3-simplex with vertices at (0,0,0), (1,0,0), (0,1,0), (0,0,1) and generates uniform points using the Dirichlet distribution method.

The LHS values being quite smaller than 1 suggests the theoretical bound is quite conservative.



